---
title: "Hello World"
date: 2019-05-07
categories: ["R"]
---

## Quarto

This is my first post! This website is developed with Quarto and served via Netlify. I tried hosting this on GitHub Pages, but had too much difficulty customizing. Having used Blogdown, I'm making the transition to Quarto.

So far this is great. I can embed R code like this:

```{r cars}
summary(cars)
```

Or include plots:

```{r sky}
pie(
  c(280, 60, 20),
  c('Sky', 'Sunny side of pyramid', 'Shady side of pyramid'),
  col = c('#0292D8', '#F7EA39', '#C4B632'),
  init.angle = -50, border = NA
)
```

------------------------------------------------------------------------

I don't anticipate blogging a lot, but I hope to make this a platform to showcase some of my pet projects. Fair warning, a lot of my projects revolve on my dog, Harper, so I literally mean "pet". Below are a few examples I hope to post about soon:

-   Raspberry Pi timelapse video of Harper crate-training
-   Harper/Not Harper image classification mobile app
-   Harper autoencoder to generate new, "fake" Harper images

Thanks for reading!
